---
title: "Cyclistic Bike-Share Program"
author: "Salah Mohamed"
date: "1/17/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Business Task Statement
How to Convert Casual riders into annual members?

## Question Asked
How do annual members and casual riders use bikes differently?

## Key Stakeholders:
1) Primary: Lily Moreno My Direct Manager

2) Secondary: Marketing Analytics Team

3) Upper Management: Cyclistic Executive team

## Goal of analysis:
To understand how casual riders and annual members use Cyclistic bikes differently

## Data Resources Used:
Trip data for both casual and annual riders made available by Motivate International Inc<https://divvy-tripdata.s3.amazonaws.com/index.html>. under this license<https://ride.divvybikes.com/data-license-agreement>.)

# Project Phases: 
The main question is How do annual members and casual riders use Cyclistic bikes differently?

## Ask:
### Deliverable:
A clear statement of the business task (How to Convert Casual riders into annual members?)

## Prepare: 
### Deliverable:

A description of all data sources used: 

1- Data is downloaded from Motivate International Inc.

2- Downloaded files trip_data_ from 2019 Q1 to 2020 Q1

3- CSV files downloaded and saved in one main csv file called(cyclistic_trip_data_2019_q1_to_q4_to_2020_q1)

4- Csv file names are (Divvy_Trips_2019_Q1.csv, Divvy_Trips_2019_Q2.csv, Divvy_Trips_2019_Q3.csv,  Divvy_Trips_2019_Q4.csv, Divvy_Trips_2020_Q1.csv)

5- Files were too large to open with excel and spreadsheets 

6- all files are imported to R in as vectors (the reason we use R because it's faster, allows me to use viz to check bias, and I want to go further to the machine learning field so it will allow me to do so because it's close to Python, and allow algorithms and machine learning)

## Process:
### Deliverable:
Documentation of any cleaning or manipulation of data, they are as follows:

Setting Working Directory
```{r}
setwd("D:/00-Data/02-Google Data Analytics/0000-Case Study/01-First Track Casestudy 1/Bikeshare Data/cyclistic_trip_data 2019_to_2020/rdirectory/00-casestudy final directory")
```

Installing and loading the packages
```{r}
install.packages("tidyverse", repos = "http://cran.us.r-project.org")
install.packages("ggplot2", repos = "http://cran.us.r-project.org")
install.packages("lubridate", repos = "http://cran.us.r-project.org")
install.packages("here", repos = "http://cran.us.r-project.org")
install.packages("skimr", repos = "http://cran.us.r-project.org")
install.packages("janitor", repos = "http://cran.us.r-project.org")
install.packages("readr", repos = "http://cran.us.r-project.org")
install.packages("Tmisc", repos = "http://cran.us.r-project.org")
```

Loading the packages
```{r}
library(lubridate)
library(tidyverse)
library(ggplot2)
library(here)
library(skimr)
library(janitor)
library(dplyr)
library(readr)
library(Tmisc)
```

Loading the data set csv files to data frames
```{r}
q1_2019 <- read_csv("Divvy_Trips_2019_Q1.csv")
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```

Quick look inside each vector to know what is up with the data
```{r}
colnames(q1_2019)
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
colnames(q1_2020)
head(q1_2019)
head(q2_2019)
head(q3_2019)
head(q4_2019)
head(q1_2020)
# Checking for duplicate values
sum(duplicated(q1_2019))
sum(duplicated(q2_2019))
sum(duplicated(q3_2019))
sum(duplicated(q4_2019))
sum(duplicated(q1_2020))
# Checking for missing values
sum(is.na(q1_2019))
sum(is.na(q2_2019))
sum(is.na(q3_2019))
sum(is.na(q4_2019))
sum(is.na(q1_2020))

glimpse(q1_2019)
glimpse(q2_2019)
glimpse(q3_2019)
glimpse(q4_2019)
glimpse(q1_2020)

# Checking the data types in each column in all data sets
str(q1_2019)
str(q2_2019)
str(q3_2019)
str(q4_2019)
str(q1_2020)
```


## Quality & Tidiness Issues Found: 
Column names are not consistent (q2_2019 has different column names need to be changed).

The trip_id column Must be al as characters.

There are some columns in (q1_2020) needs to be removed (rideable_type, start_lat, start_lng, end_lat, end_lng).

There are some columns in (q1_2019, q2_2019, q3_2019, q4_2019) to remove (gender, birthyear).

No Duplicates Found.

There are missing values (q1_2019, gender, birthyear), (q2_2019, birthyear), (q3_2019, gender birthyear),(q4_2019, gender birthyear), (q1_2020, 4 values).


### First Cleaning Phase
```{r}
# Renaming all columns into specified consistent names
q1_2019 <- rename(q1_2019
                  ,trip_id = trip_id
                  ,start_time = start_time
                  ,end_time = end_time
                  ,bike_id = bikeid
                  ,trip_duration = tripduration
                  ,start_station_id = from_station_id
                  ,start_station_name = from_station_name
                  ,end_station_id = to_station_id
                  ,end_station_name = to_station_name
                  ,user_type = usertype
                  ,gender = gender
                  ,birth_year = birthyear )

q2_2019 <- rename(q2_2019
                  ,trip_id = "01 - Rental Details Rental ID"
                  ,start_time = "01 - Rental Details Local Start Time" 
                  ,end_time = "01 - Rental Details Local End Time"
                  ,bike_id = "01 - Rental Details Bike ID"
                  ,trip_duration = "01 - Rental Details Duration In Seconds Uncapped" 
                  ,start_station_id = "03 - Rental Start Station ID"
                  ,start_station_name = "03 - Rental Start Station Name"
                  ,end_station_id = "02 - Rental End Station ID"
                  ,end_station_name = "02 - Rental End Station Name"
                  ,user_type = "User Type"
                  ,gender = "Member Gender"
                  ,birth_year = "05 - Member Details Member Birthday Year")

q3_2019 <- rename(q3_2019
                  ,trip_id = trip_id
                  ,start_time = start_time
                  ,end_time = end_time
                  ,bike_id = bikeid
                  ,trip_duration = tripduration
                  ,start_station_id = from_station_id
                  ,start_station_name = from_station_name
                  ,end_station_id = to_station_id
                  ,end_station_name = to_station_name
                  ,user_type = usertype
                  ,gender = gender
                  ,birth_year = birthyear)

q4_2019 <- rename(q4_2019
                  ,trip_id = trip_id
                  ,start_time = start_time
                  ,end_time = end_time
                  ,bike_id = bikeid
                  ,trip_duration = tripduration
                  ,start_station_id = from_station_id 
                  ,start_station_name = from_station_name
                  ,end_station_id = to_station_id
                  ,end_station_name = to_station_name
                  ,user_type = usertype
                  ,gender = gender
                  ,birth_year = birthyear)

q1_2020 <- rename(q1_2020
                  ,trip_id = ride_id
                  ,start_time = started_at
                  ,end_time = ended_at
                  ,bike_id = rideable_type
                  ,start_station_id = start_station_id 
                  ,start_station_name = start_station_name
                  ,end_station_id = end_station_id
                  ,end_station_name = end_station_name
                  ,user_type = member_casual)

# Convert ride_id in all dataframes to characters to allow merging all in one data set 
q1_2019 <-  mutate(q1_2019, trip_id = as.character(trip_id))
q2_2019 <-  mutate(q2_2019, trip_id = as.character(trip_id))
q3_2019 <-  mutate(q3_2019, trip_id = as.character(trip_id))
q4_2019 <-  mutate(q4_2019, trip_id = as.character(trip_id))
q1_2020 <-  mutate(q1_2020, trip_id = as.character(trip_id))
q1_2019 <-  mutate(q1_2019, bike_id = as.character(bike_id))
q2_2019 <-  mutate(q2_2019, bike_id = as.character(bike_id))
q3_2019 <-  mutate(q3_2019, bike_id = as.character(bike_id))
q4_2019 <-  mutate(q4_2019, bike_id = as.character(bike_id))
q1_2020 <-  mutate(q1_2020, bike_id = as.character(bike_id))
```

Merging individual quarter's into one main data frame
```{r}
trip_data <- bind_rows(q1_2019, q2_2019, q3_2019, q4_2019, q1_2020)
```

### Assessing all_trips data frame to stand on it's quality and tidiness.
```{r}
str(trip_data)
glimpse(trip_data)
summary(trip_data)
class(trip_data)
class(trip_data$start_time)
class(trip_data$end_time)
class(trip_data$bike_id)
class(trip_data$trip_duration)
class(trip_data$start_station_name)
class(trip_data$end_station_name)
class(trip_data$user_type)
class(trip_data$birth_year)
```

## More issues found with the new full data frame

Start and end time are a group of datetime needs to be separated to YY MM DD time.

There are unneeded columns need to be removed (bike_id, start_station_id, end_station_id, gender, start_lat, start_lng, end_lat, end_lng).


### Second cleaning phase
```{r}
# Creating a date column to be able to explore trends based on the day or month, then column for YY MM DD Dayof week
trip_data$date <- as.Date(trip_data$start_time)
trip_data$month <- format(as.Date(trip_data$date), "%M")
trip_data$day <- format(as.Date(trip_data$date), "%D")
trip_data$year <- format(as.Date(trip_data$date), "%Y")
trip_data$day_of_week <- format(as.Date(trip_data$date), "%A")

# Removing unneeded variables
trip_data <- trip_data %>%  
  select(-c(start_station_id, end_station_id, gender, start_lat, start_lng, end_lat, end_lng))
# Rechecking the data frame
glimpse(trip_data)
str(trip_data)
```

Assessing the quality & Tidiness of each variable
```{r}
names(trip_data)
unique(trip_data$start_station_name)
unique(trip_data$end_station_name)
unique(trip_data$user_type)
unique(trip_data$birth_year)
unique(trip_data$date)
unique(trip_data$trip_duration)
unique(trip_data$day_of_week)
sum(is.na(trip_data$trip_duration))
```

# More Issues found
start_station name includes unneeded values("MTL-ECO5.1-01", "HQ QR" )

End station name also has unneeded values(HQ QR", "MTL-ECO5.1-01" ) and NA values

User type column values need to be consistent (member& casual only)

Week days are not ordered correctly

There are NA's in trip duration column so we will calculate it manually using start and end time

### Final Cleaning
```{r}
# replacing "Subscriber" with "member" and "Customer" with "casual"
trip_data <-  trip_data %>% 
  mutate(user_type = recode(user_type
                            ,"Subscriber" = "member"
                            ,"Customer" = "casual"))

# Adding a "ride_length" calculation to all_trips (in seconds)
trip_data$ride_length <- difftime(trip_data$end_time,trip_data$start_time)

# Converting ride length to minutes
trip_data <- trip_data %>%
  mutate(ride_length_minu=ride_length/60)

# Convert "ride_length" from Factor to numeric so we can run calculations on the data
trip_data$ride_length_minu <- as.numeric(trip_data$ride_length_minu)

# Ordering days of week properly
trip_data$day_of_week <- ordered(trip_data$day_of_week, levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

# Removing "bad" data
trip_data <- trip_data[!(trip_data$start_station_name == "HQ QR" | trip_data$ride_length<0),]
```

### Final Check
```{r}
colnames(trip_data)  #List of final column names
nrow(trip_data)  #How many rows are in the final data frame?
dim(trip_data)  #Dimensions of the final data frame?
head(trip_data, 20)  # the first 20 rows of the final data frame.  
tail(trip_data, 20 ) # the last 20 rows of the final data frame.
str(trip_data)       # list of columns and data types (numeric, character, etc)
summary(trip_data)  # Basic Statistical summary of data
```

Rechecking the most important variables in the Data which will be guiding during the analysis phase
```{r}
unique(trip_data$start_station_name)
unique(trip_data$user_type)
unique(trip_data$birth_year)
unique(trip_data$date)
unique(trip_data$ride_length_minu)
unique(trip_data$day_of_week)
sum(is.na(trip_data$trip_duration))
```

## Analyze:
### Key tasks

Aggregate your data so it’s useful and accessible.

Organize and format your data.

Perform calculations.

Identify trends and relationships.

### Deliverable:
A summary of your analysis

```{r}
# Descriptive analysis on ride_length in minutes
mean(trip_data$ride_length_minu) 
median(trip_data$ride_length_minu) 
max(trip_data$ride_length_minu) 
min(trip_data$ride_length_minu) 
summary(trip_data$ride_length_minu)

# Comparing ride_length per minute for members and casual users
aggregate(trip_data$ride_length_minu ~ trip_data$user_type, FUN = mean)
aggregate(trip_data$ride_length_minu ~ trip_data$user_type, FUN = median)
aggregate(trip_data$ride_length_minu ~ trip_data$user_type, FUN = min)
aggregate(trip_data$ride_length_minu ~ trip_data$user_type, FUN = max)

# comparing ride_length per minute for members and casual users during different week days and creating visuals
aggregate(trip_data$ride_length_minu ~ trip_data$user_type + trip_data$day_of_week, FUN = mean)
aggregate(trip_data$ride_length_minu ~ trip_data$user_type + trip_data$day_of_week, FUN = min)
aggregate(trip_data$ride_length_minu ~ trip_data$user_type + trip_data$day_of_week, FUN = max)

# The most used start station in general and  per each user type
count(trip_data, start_station_name, sort = TRUE)
count(trip_data, end_station_name, sort = TRUE)
count(trip_data, start_station_name, sort = FALSE)
count(trip_data, end_station_name, sort = FALSE)
count(trip_data, user_type, start_station_name,sort= TRUE)
count(trip_data, start_station_name, user_type, sort= FALSE)
count(trip_data, user_type, day_of_week, sort = TRUE)
count(trip_data, user_type, sort = FALSE)
names(trip_data)
```

## Share:

Share is divided by two steps Key insights found from conduction some descriptive analysis, then a final file version is made and saved as CSV file to create visuals using Tableau.


## Key Insights & surprises about the data:

1- The average ride length per minute for casual users is more than premium users.

2- Friday has the longest avg_ride_length_per_minute for casual users, However Saturday has the longest avg ride per minute for annual members.

3- The maximum ride length per minute for member users occurs on Tuesday, however the maximum ride length per minute for casual users occur on Thursday.

4- The most used start station in general is "Streeter Dr & Grand Ave " with the amount of 70447 occurrences, Also which surprising it's the most used end station name by 84479 occurrences

5- The least used start_station_name is 2112 W Peterson Ave, with 546 occurrences, and surprisingly it's the same least used end station with 685.

6- The most used station for annual members is Canal St & Adams St 58161, and for casual users Streeter Dr & Grand Ave with 54634

7- The least used start station is 2112 W Peterson Ave for both types of users.

8- Total number of rides for the period of (2019-01-01 to 2020-03-30) is 4,241,111rides, Noticed that Annual members have 3315766 rides with the % of 78.18%, and casual members have 925345 rides with the % of 21.81%.

9- The top week day for casual riders bike rides is Saturday, and for annual members is Tuesday.

10- The lowest rides for casual riders is Tuesday, and for annual members is Sunday.


My Recommendations:
1- We can start our campaign by conducting a survey for casual riders on Saturday since it's the top day for casual users.

2- we can conduct a marketing campaign on Saturday, and Streeter Dr & Grand Ave since it's the top for casual users
3- finally we can make a discount for users only on Thursday, since it's the day when most casual riders have the longest avg ride_length
## Creating a Trip_data_for_visualizing file
```{r}
final_trip_data <- trip_data %>% drop_na() %>% 
  select(-c(start_time, end_time, trip_duration, ride_length, year))

write.csv(final_trip_data, 
          file = 'D:/00-Data/02-Google Data Analytics/0000-Case Study/01-First Track Casestudy 1/Bikeshare Data/cyclistic_trip_data 2019_to_2020/rdirectory/00-casestudy final directory/final_trip_data.csv')

```

